Code used to generate paper results : Database comparison and CT characteristics

### 2023-08-23 Paper_code-Database_CT.Rmd
Separated full code in two files

---
title: "Code used to generate paper results"
author: "Julie Devis"
date: "2023-08-28"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

@@

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(Vennerable)
library(biomaRt)
library(tidyverse)
library(SummarizedExperiment)
library(factoextra)
library(UpSetR)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(org.Hs.eg.db)
library(clusterProfiler)
library(msigdbr)
library(DOSE)
library(tibble)
library(BiocParallel)
library(patchwork)
library(Biostrings)
```

@@

# Package application : What are the mechanisms involved in CT demethylation in cancer

## Determination of PMDs in male fibroblasts celles

WGBS data from GM23248 cells (male so X chromosome is reliable) downloaded from 
ENCODE.
Data was processed using a WGBS pipeline [MethPipe]
(http://smithlabresearch.org/software/methpipe/) including [TrimGalore](https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/) 
to remove low quality reads and trim the adapter from the sequences.

The pipeline has been verified by comparing PMDs already defined by Lister et 
al. on IMR90 and PMDs obtained with MethPipe.

```{bash download_GM23248, eval = FALSE}
wget https://www.encodeproject.org/files/ENCFF142GZA/@@download/ENCFF142GZA.fastq.gz
wget https://www.encodeproject.org/files/ENCFF845RUF/@@download/ENCFF845RUF.fastq.gz
wget https://www.encodeproject.org/files/ENCFF968GRR/@@download/ENCFF968GRR.fastq.gz
wget https://www.encodeproject.org/files/ENCFF975YNS/@@download/ENCFF975YNS.fastq.gz
```

```{bash trim/methpipe_GM23248, eval = FALSE}
#!/bin/bash
mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV;
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV
/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF142GZA.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF845RUF.fastq

mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI;
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI
/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF975YNS.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF968GRR.fastq

#!/bin/bash

cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep1_ENCLB777FMV;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 ENCFF142GZA_val_1.fq \
-2 ENCFF845RUF_val_2.fq \
-o rep1_ENCLB777FMV_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep1_ENCLB777FMV.mr.sorted_start rep1_ENCLB777FMV_mapping.mr;
/opt/methpipe/build/duplicate-remover -S rep1_ENCLB777FMV_dremove_stat.txt -o rep1_ENCLB777FMV.mr.dremove rep1_ENCLB777FMV.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep1_ENCLB777FMV.bsrate rep1_ENCLB777FMV.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep1_ENCLB777FMV.meth rep1_ENCLB777FMV.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o rep1_ENCLB777FMV_CpG_merge.meth rep1_ENCLB777FMV.meth;

#Stats
/opt/methpipe/build/levels -o rep1_ENCLB777FMV.levels rep1_ENCLB777FMV.meth;

cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep2_ENCLB436YZI;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 ENCFF975YNS_val_1.fq \
-2 ENCFF968GRR_val_2.fq \
-o rep2_ENCLB436YZI_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep2_ENCLB436YZI.mr.sorted_start rep2_ENCLB436YZI_mapping.mr;
/opt/methpipe/build/duplicate-remover -S rep2_ENCLB436YZI_dremove_stat.txt -o rep2_ENCLB436YZI.mr.dremove rep2_ENCLB436YZI.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep2_ENCLB436YZI.bsrate rep2_ENCLB436YZI.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep2_ENCLB436YZI.meth rep2_ENCLB436YZI.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o rep2_ENCLB436YZI_CpG_merge.meth rep2_ENCLB436YZI.meth;

#Stats
/opt/methpipe/build/levels -o rep2_ENCLB436YZI.levels rep2_ENCLB436YZI.meth;

#Merge both replicats
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248;

/opt/methpipe/build/merge-methcounts rep1_ENCLB777FMV/rep1_ENCLB777FMV_CpG_merge.meth \
rep2_ENCLB436YZI/rep2_ENCLB436YZI_CpG_merge.meth \
-o GM23248_merged.meth;

#pmd
/opt/methpipe/build/pmd -v -i 1000 -o GM23248.pmd GM23248_merged.meth
```

```{r import_meth_GM23248, eval = FALSE}
GM23248_meth <- read_delim("GM23248_merged.meth", "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(GM23248_meth) <- c("chromosome", "position", "strand", "sequence_context","methylation_level","number_of_reads")
```

We can use the PMD file coming from MethPipe and change it into a GRanges object
so that we can easily do overlapping analysis

```{r load_PMD, message = FALSE}
my_PMD_GM23248 <- read_delim("data/GM23248.pmd", "\t", 
                             escape_double = FALSE, col_names = FALSE, 
                             trim_ws = TRUE)
names(my_PMD_GM23248) <- c("chr", "start","end","name_quality","bins","strand")
my_PMD_GM23248 <- separate(my_PMD_GM23248, "name_quality", 
                           into = c("name", "likelihood_start", "certainty_start",
                                    "likelihood_end", "certainty_end"), 
                           sep = ":")
my_PMD_GM23248_GR <- makeGRangesFromDataFrame(my_PMD_GM23248 [, -10], keep.extra.columns = TRUE)
```

Not PMD and NA regions aren’t dissociated. Coverage of windows of 10kb were 
calculated along the genome. A threshold of 1000 reads per 10kb is necessary to 
not be considered NA.

```{r all_chr_10kb, eval = FALSE}
for (chromosome_name in paste0("chr",c(1:22,"X","Y"))){

  roi <- GM23248_meth %>% 
    filter(chromosome == chromosome_name)

  res <- data.frame(region_i = floor(min(roi$position)/10000):floor(max(roi$position)/10000),
                    reads_tot = rep(NA,length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))),
                    chromosome = rep(unique(roi$chromosome),length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))))

  res <- res %>% 
    mutate(start_region = region_i*10000) %>% 
    mutate(end_region = region_i*10000+10000) %>% 
    column_to_rownames("region_i")

  for (i in floor(min(roi$position)/10000):floor(max(roi$position)/10000)){
    value <-  i*10000
    tmp <- roi %>% 
      filter(position >= value & position < (value + 10000))
    res[as.character(i) , "reads_tot"] <- sum(tmp$number_of_reads)
    }

  res_GR <- makeGRangesFromDataFrame(res, start.field = "start_region", 
                                     end.field = "end_region", keep.extra.columns = TRUE)

  overlap_boul <- overlapsAny(res_GR, my_PMD_GM23248_GR, type = "within")

  res$in_PMD <- overlap_boul

  saveRDS(res, file = paste0("data/10kb_reads_PMD_", chromosome_name, ".rds"))

}

windows_reads_PMD_chr1 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr1.rds")
windows_reads_PMD_chr2 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr2.rds")
windows_reads_PMD_chr3 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr3.rds")
windows_reads_PMD_chr4 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr4.rds")
windows_reads_PMD_chr5 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr5.rds")
windows_reads_PMD_chr6 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr6.rds")
windows_reads_PMD_chr7 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr7.rds")
windows_reads_PMD_chr8 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr8.rds")
windows_reads_PMD_chr9 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr9.rds")
windows_reads_PMD_chr10 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr10.rds")
windows_reads_PMD_chr11 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr11.rds")
windows_reads_PMD_chr12 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr12.rds")
windows_reads_PMD_chr13 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr13.rds")
windows_reads_PMD_chr14 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr14.rds")
windows_reads_PMD_chr15 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr15.rds")
windows_reads_PMD_chr16 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr16.rds")
windows_reads_PMD_chr17 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr17.rds")
windows_reads_PMD_chr18 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr18.rds")
windows_reads_PMD_chr19 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr19.rds")
windows_reads_PMD_chr20 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr20.rds")
windows_reads_PMD_chr21 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr21.rds")
windows_reads_PMD_chr22 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr22.rds")
windows_reads_PMD_chrX <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrX.rds")
windows_reads_PMD_chrY <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrY.rds")

windows_reads_PMD <- rbind(windows_reads_PMD_chr1, windows_reads_PMD_chr2, windows_reads_PMD_chr3, windows_reads_PMD_chr4, windows_reads_PMD_chr5, windows_reads_PMD_chr6, windows_reads_PMD_chr7, windows_reads_PMD_chr8, windows_reads_PMD_chr9, windows_reads_PMD_chr10, windows_reads_PMD_chr11, windows_reads_PMD_chr12, windows_reads_PMD_chr13, windows_reads_PMD_chr14, windows_reads_PMD_chr15, windows_reads_PMD_chr16, windows_reads_PMD_chr17, windows_reads_PMD_chr18, windows_reads_PMD_chr19, windows_reads_PMD_chr20, windows_reads_PMD_chr21, windows_reads_PMD_chr22, windows_reads_PMD_chrX, windows_reads_PMD_chrY)

rownames(windows_reads_PMD) <- NULL

saveRDS(windows_reads_PMD, file = "data/10kb_reads_PMD_all.rds")
```

Here we load the coverage for all the 10k windows and determine which one are 
non-PMD due to coverage issues. These region are to be removed from the analysis.

```{r NA_windows}
windows_reads_PMD <- readRDS("data/10kb_reads_PMD_all.rds")

NA_nonPMD_GR <- windows_reads_PMD %>% 
  filter(!in_PMD) %>% 
  filter(reads_tot <= 1000) %>% 
  makeGRangesFromDataFrame(start.field = "start_region", end.field = "end_region",
                           keep.extra.columns = TRUE) %>% 
  GenomicRanges::reduce()
```

## Are CT genes associated with PMDs ?

### CT regulated by methylation and PMD overlap

```{r CT_genes_prep}
CT_genes <- CT_genes %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000))

CT_genes$strand <- ifelse(CT_genes$strand == "1", "+", "-")
seqlevelsStyle(CT_genes$chr) <- "UCSC"

CT_meth_regulated <- CT_genes %>% 
  filter(regulated_by_methylation)
```

```{r CT_prom}
meth_reg_prom_GR <- makeGRangesFromDataFrame(CT_meth_regulated, start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)
```

```{r overlap_PMD_CT}
CTmeth_PMD_GR <- subsetByOverlaps(meth_reg_prom_GR, my_PMD_GM23248_GR)
CTmeth_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
                                                       my_PMD_GM23248_GR, 
                                                   invert = TRUE),
                                      NA_nonPMD_GR, invert = TRUE)
CTmeth_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
                                                     my_PMD_GM23248_GR, 
                                                 invert = TRUE), 
                                    NA_nonPMD_GR)
```

Now that these overlapping have been done, I can report them into my data as a 
boulean column to use analyse it.

```{r PMD_in_CT}
CT_meth_regulated <- CT_meth_regulated %>% 
  mutate(PMD = case_when(CT_meth_regulated$external_gene_name %in% CTmeth_PMD_GR$external_gene_name ~ TRUE,
                         CT_meth_regulated$external_gene_name %in% CTmeth_not_PMD_GR$external_gene_name ~ FALSE,
                         CT_meth_regulated$external_gene_name %in% CTmeth_NA_PMD_GR$external_gene_name ~ NA))
table(is.na(CT_meth_regulated$PMD))
table(CT_meth_regulated$PMD)

PMD_perc <- table(CT_meth_regulated$PMD)[[2]] / 
  sum(table(CT_meth_regulated$PMD))*100
PMD_perc

X_in_PMD_table <- CT_meth_regulated %>% 
  filter(X_linked) %>% 
  dplyr::select(PMD) %>% 
  table()
X_in_PMD_table

PMD_chr_repartition <- CT_meth_regulated %>% 
  filter(!is.na(PMD)) %>% 
  dplyr::select(X_linked) %>% 
  table()
PMD_chr_repartition
```

### CT not regulated by methylation and PMD overlap

```{r CT_not_meth_PMD}
CT_not_meth_regulated <- CT_genes %>% 
  filter(!regulated_by_methylation)

prom_GR_not_reg <- makeGRangesFromDataFrame(CT_not_meth_regulated, 
                                            start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)
not_reg_PMD_GR <- subsetByOverlaps(prom_GR_not_reg, my_PMD_GM23248_GR)
not_reg_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
                                                        my_PMD_GM23248_GR, 
                                                   invert = TRUE), 
                                  NA_nonPMD_GR, invert = TRUE)
not_reg_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
                                                      my_PMD_GM23248_GR, 
                                                 invert = TRUE),
                                NA_nonPMD_GR)

CT_not_meth_regulated <- CT_not_meth_regulated %>% 
  mutate(PMD = case_when(CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_PMD_GR$external_gene_name ~ TRUE,
                         CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_not_PMD_GR$external_gene_name ~ FALSE,
                         CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_NA_PMD_GR$external_gene_name ~ NA))

table(is.na(CT_not_meth_regulated$PMD))
table(CT_not_meth_regulated$PMD)

CT_not_meth_reg_PMD_perc <- table(CT_not_meth_regulated$PMD)[[2]] / 
  sum(table(CT_not_meth_regulated$PMD))*100
CT_not_meth_reg_PMD_perc
```

### Enrichment analysis

The numbers are useful but we don't know if it is significant or not. 
In order to do so, we're going to compare this percentage with an empiric 
distribution. For that we need groups of genes comparable.

This distribution is going to be realised by 1000 times computing the % of genes
in PMD in `r PMD_chr_repartition[[1]]` tissue-specific genes on autosomes and 
`r PMD_chr_repartition[[2]]` tissue-specific genes on the chromosome X. We use 
tissue-specific genes because this characteristic can make them more prone to be
in a repressive region like PMDs.

We then have to create a pool of non-CT tissue-specific genes with their 
localisation, for that we use the list of tissue specific genes that can be 
found on the Human Protein Atlas (3106 tissue enriched genes: At least four-fold
higher mRNA level in a particular tissue compared to any other tissue).

It is necessary to use biomaRt because we need transcription start sites to be 
able to realise the same analysis on enriched genes than on CT genes. Also, we 
need to remove CT genes and genes that are in the non-covered regions.

```{r enriched_genes}
enriched_genes <- read.delim("data/HPA_tissue_enriched_genes.tsv")

attributes_vector <- c("ensembl_gene_id","external_gene_name", "chromosome_name",
                       "strand", "transcription_start_site", 
                       "external_transcript_name", "transcript_is_canonical")
biomaRt_query <- getBM(attributes = attributes_vector, mart = ensembl)
enriched_genes_biomart <- biomaRt_query %>% 
  filter(ensembl_gene_id %in% enriched_genes$Ensembl)
enriched_genes_biomart <- enriched_genes_biomart %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000)) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  filter(transcript_is_canonical == 1) %>%
  filter(chromosome_name%in%c(1:22, "X", "Y"))

enriched_genes_biomart$strand <- ifelse(enriched_genes_biomart$strand == "1", "+", "-")
seqlevelsStyle(enriched_genes_biomart$chromosome_name) <- "UCSC"

enriched_genes_GR <- makeGRangesFromDataFrame(enriched_genes_biomart, 
                                              start.field = "start_prom",
                                              end.field = "end_prom",
                                              keep.extra.columns = TRUE)
enriched_NA_GR <-subsetByOverlaps(subsetByOverlaps(enriched_genes_GR, 
                                                   my_PMD_GM23248_GR, 
                                                   invert = TRUE),
                                  NA_nonPMD_GR)

enriched_genes_analysable <- as.data.frame(
  enriched_genes_GR[!enriched_genes_GR$ensembl_gene_id %in% enriched_NA_GR$ensembl_gene_id, ])
```

```{r enrichment_analysis}
random_percentage_chrprop <- function(i, data, nchrX, nautosomes){
  random_sample1 <- sample_n(filter(data, seqnames == "chrX"), nchrX)
  random_sample2 <- sample_n(filter(data, seqnames != "chrX"), nautosomes)
  random_sample <- rbind(random_sample1,random_sample2)
  random_sample_GR <- makeGRangesFromDataFrame(random_sample, 
                                               keep.extra.columns = TRUE)
  length(subsetByOverlaps(random_sample_GR, my_PMD_GM23248_GR)) / 
    length(random_sample_GR)*100
}

distrib_enriched <- unlist(bplapply(1:1000, random_percentage_chrprop, 
                                    data = enriched_genes_analysable, 
                                    nchrX = PMD_chr_repartition[[2]], 
                                    nautosomes = PMD_chr_repartition[[1]]))
hist(distrib_enriched,
     col="violetred1",
     border = "violetred3",
     breaks = 20,
     main = paste0("Percentage of promoters in PMDs when randomly taking ",
                   sum(table(CT_meth_regulated$PMD)), 
                   " tissue-specific 
genes 1000 times (taking chromosome proportion into account)"),
     cex.main = 0.75,
     xlab = "Percentages",
     ylab = "Frequency",
     xlim = c(20,60))
abline(v = PMD_perc, col = "slateblue1", lwd = 3)
text(PMD_perc, 120, "CT genes regulated 
by methylation", col = "slateblue4", cex = 0.8)
abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
text(CT_not_meth_reg_PMD_perc, 120, "CT genes not regulated 
by methylation", col = "darkseagreen", cex = 0.8)

```{r add_PMD}
PMD_CT_results <- rbind(dplyr::select(CT_not_meth_regulated, 
                                      c(PMD, external_gene_name)),
                    dplyr::select(CT_meth_regulated, 
                                  c(PMD, external_gene_name)))
CT_genes <- left_join(CT_genes, PMD_CT_results)
```

## Methylation in an aging model : CD4+ cells

Now that we know that CT genes are indeed enriched in PMDs, we have to go 
further. Indeed, we know that proliferation has an impact on PMD and their 
methylation : there are losses of methylation with time. As CT genes are located
in PMDs, we can hypothesise that their methylation might be impacted by the 
losses happening during proliferation. We thus want to focus on the impact on CT
genes of the methylation changes happening on PMDss in proliferative or aging 
cells.

### Data processing

Data coming from Heyn 2012 paper.

```{bash CD4, eval = FALSE}
#!/bin/bash

#Newborn
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330578
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330579
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394135
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394136
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394137
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394138
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394139
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394140
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394141
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394142

#Centenarian
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330574
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330576
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330577
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330575

#Middle-age
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389249
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389248

for f in `cat CD4_SRA.txt`;

do mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f; 
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f;

#Trimming
/opt/TrimGalore/trim_galore --fastqc --path_to_cutadapt /home/users/jdevis/cutadapt --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_1.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_2.fastq;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 $f\_1_val_1.fq \
-2 $f\_2_val_2.fq \
-o $f\_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o $f.mr.sorted_start $f\_mapping.mr;
/opt/methpipe/build/duplicate-remover -S $f\_dremove_stat.txt -o $f.mr.dremove $f.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o $f.bsrate $f.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o $f.meth $f.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o $f\_CpG_merge.meth $f.meth;

#Stats
/opt/methpipe/build/levels -o $f.levels $f.meth

done

Bisulfite-Seq data coming from a male newborn, a middle age donor and a male 
centenarian.

Quality of the data is good, with a mean of 13 reads per cytosine. However, 
middle age data has no info about the sex of the donor. By comparing the number 
of reads by chromosome in this sample, a female one and the male ones, we saw 
that the proportion of reads on chromosomes X and Y showed that the middle age 
data come from a woman. We then have to be very careful when analysing the X 
chromosome with the middle age data. To facilitate the analysis and compare all
samples, I’ll remove the adult data on chrX.

```{r load_CD4, message = FALSE, warning = FALSE}
columns <- c("chromosome", "position", "strand", "sequence context","methylation_level","number_of_reads")
CD4_newborn <- read_delim("data/CD4_newborn.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_newborn) <- columns
CD4_middle <- read_delim("data/CD4_middle.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_middle) <- columns
CD4_middle <- filter(CD4_middle, chromosome != "chrX")
CD4_cent <- read_delim("data/CD4_cent.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_cent) <- columns
```
```{r change_CD4, message = FALSE, warning = FALSE}
CD4_newborn$age <- "newborn"
CD4_middle$age <- "middle_age"
CD4_cent$age <- "centenarian"
```
```{r CD4_GR, message = FALSE, warning = FALSE}
CD4_newborn_GR <- makeGRangesFromDataFrame(CD4_newborn, start.field = "position", 
                                           end.field = "position", keep.extra.columns = TRUE)
CD4_middle_GR <- makeGRangesFromDataFrame(CD4_middle, start.field = "position", 
                                          end.field = "position", keep.extra.columns = TRUE)
CD4_cent_GR <- makeGRangesFromDataFrame(CD4_cent, start.field = "position", 
                                        end.field = "position", keep.extra.columns = TRUE)
```
```{r rm, echo = FALSE}
rm(CD4_newborn)
rm(CD4_middle)
rm(CD4_cent)
```

### CT genes methylation level

Here I extract methylation for all CpGs located in CT genes promoter 
(regulated by methylation because it's these that we can find in PMDs).

```{r CT_CD4, warning = FALSE}
CT_prom_GR <- makeGRangesFromDataFrame(CT_genes, start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)

newborn <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
newborn$external_gene_name <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, 
                                              ignore.strand = TRUE)$external_gene_name
newborn <- as.data.frame(newborn)
middle <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
middle$external_gene_name <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, 
                                             ignore.strand = TRUE)$external_gene_name
middle <- as.data.frame(middle)
cent <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
cent$external_gene_name <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, 
                                           ignore.strand = TRUE)$external_gene_name
cent <- as.data.frame(cent)

CTmeth_inCD4 <- rbind(newborn, middle, cent)
rm(newborn, middle, cent)
CTmeth_inCD4$age <- factor(CTmeth_inCD4$age, 
                           levels = c("newborn","middle_age","centenarian"))
CTmeth_inCD4[CTmeth_inCD4$number_of_reads == 0,]$methylation_level <- NA

length(unique(CTmeth_inCD4$external_gene_name))

We can see here that we have information for all 
`r length(unique(CTmeth_inCD4$external_gene_name))` CT genes. We now have to 
calculate the mean methylation level of each promoter to visualise it and 
compare to PMDs.

Some df that we obtain are empty, their CpG are not covered by the CD4 data, 
and so need to be removed. I also want to remove those who have data in only 
one age category, as we want to compare different age, they are of no need. 
When one of the age is missing, I'm adding it as NA to facilitate the rest 
(mostly gene on chrX and we removed data from adult)

```{r CT_level}
reorganize_data_genes <- function(gene_name, data){
  reorganized_data <- data %>%
    filter(external_gene_name == gene_name) %>% 
    dplyr::select("start","methylation_level","age") %>% 
    na.omit() %>%  # to remove CpG rows with NA meth_level
    pivot_wider(names_from = start, values_from = methylation_level) %>% 
    column_to_rownames("age")
}

CT_per_gene_CD4_datasets <- sapply(unique(CTmeth_inCD4$external_gene_name), 
                                   reorganize_data_genes, data = CTmeth_inCD4)

p <- rep(NA, length(CT_per_gene_CD4_datasets))
for(i in 1:length(CT_per_gene_CD4_datasets)){
 p[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) > 1 & 
   is.data.frame(CT_per_gene_CD4_datasets[[i]])
}
CT_per_gene_CD4_datasets <- CT_per_gene_CD4_datasets[p]

k <- rep(NA, length(CT_per_gene_CD4_datasets))
for(i in 1:length(CT_per_gene_CD4_datasets)){
  k[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) == 2
}
for (gene in names(CT_per_gene_CD4_datasets[k])){
  tmp <- rep(NA, ncol(CT_per_gene_CD4_datasets[[gene]]))
  CT_per_gene_CD4_datasets[[gene]] <- rbind(CT_per_gene_CD4_datasets[[gene]], tmp)
  rownames(CT_per_gene_CD4_datasets[[gene]])[3] <- 
    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in% 
                                                rownames(CT_per_gene_CD4_datasets[[gene]]))]
  # naming the new row
}

for (gene in names(CT_per_gene_CD4_datasets)){
  CT_per_gene_CD4_datasets[[gene]] <- arrange(CT_per_gene_CD4_datasets[[gene]], 
                                              rownames(CT_per_gene_CD4_datasets[[gene]]))
  # Put rows in the right order
}

rm(p, i, k, tmp)

CT_meth_means_pre <- sapply(CT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)

CT_meth_means <- CT_meth_means_pre %>%
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("external_gene_name")

>>>>

---
title: "Code used to generate paper results : Database comparison and CT characteristics"
author: "Julie Devis"
date: "2023-08-28"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

@@

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(Vennerable)
library(biomaRt)
library(tidyverse)
library(SummarizedExperiment)
library(factoextra)
library(UpSetR)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(org.Hs.eg.db)
library(clusterProfiler)
library(msigdbr)
library(DOSE)
library(BiocParallel)
library(patchwork)
library(Biostrings)
```

@@
